Determinism, Complexity, and Predictability in Computer Performance

### 2014-03-28 wpe.tex
more edits to wpe

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present which is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  In theory, the
more redundancy in a signal, the more predictable it is {\color{red}
  [[citation for that theory?]]}.  Redundancy is critical to
prediction, in general.  And the specific \emph{form} of the
redundancy dictates whether a particular prediction method will work
well or poorly on the corresponding signal.  A linear method cannot
detect or make use of nonlinear redundancy, for instance.  This issue
is central to the claims in this paper and the discussion in the
following section.

@@

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
``natural complexity measure for time series''
\cite{bandt2002per}. Permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination.  Rather than looking at the
statistics of sequences of values, as is done when computing the
Shannon entropy, permutation entropy looks at the statistics of the
\emph{orderings} of sequences of values using ordinal
analysis. Ordinal analysis of a time series is the process of mapping
successive time-ordered elements of a time series to their
value-ordered permutation of the same size.  By way of example, if
$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
$321$.  {\color{red} [[This method has many features; among other
      things, it is generally robust to observational noise and
      requires no knowledge of the underlying mechanisms.]] [[This
      sentence is out of place.]]}

>>>>

From a different point of view, we can consider the information (as
measured by the Shannon entropy) contained in a single observable, say
the \emph{present}, of the system. This information can be partitioned
into two components: the information shared with past
observations---e.g., the mutual information between the past and
present---and the information in the present that is not contained in
the past (aka ``the conditional entropy of the present given the
past'').  The first part is known as the \emph{redundancy}, as it is
information in the present which is also in the past.  The second part
is the aforementioned \emph{Shannon entropy rate}.  We conjecture that 
the more redundancy in a signal, the more predictable it is.  
Redundancy is critical to prediction, in general.  And the specific 
\emph{form} of the redundancy dictates whether a particular prediction 
method will work well or poorly on the corresponding signal.  A linear 
method cannot detect or make use of nonlinear redundancy, for 
instance.  This issue is central to the claims in this paper and the 
discussion in the following section.

@@

Bandt and Pompe introduced the \emph{permutation entropy} (PE) as a
``natural complexity measure for time series''
\cite{bandt2002per}. The permutation entropy employs a method of
discretizing real-valued time series that follows the intrinsic
behavior of the system under examination, and has many advantages 
including being robust to observational noise and its application does 
not require any knowledge of the underlying mechanisms.  Rather than 
looking at the statistics of sequences of values, as is done when 
computing the Shannon entropy, permutation entropy looks at the 
statistics of the \emph{orderings} of sequences of values using 
ordinal analysis. Ordinal analysis of a time series is the process of 
mapping successive time-ordered elements of a time series to their
value-ordered permutation of the same size.  By way of example, if
$(x_1, x_2, x_3) = (9, 1, 7)$ then its \emph{ordinal pattern},
$\phi(x_1, x_2, x_3)$, is $231$ since $x_2 \leq x_3 \leq x_1$.  The
ordinal pattern of the permutation $(x_1, x_2, x_3) = (9, 7, 1)$ is
$321$.
